Virtualization of a Real-Time Operating System for Robot Control with a Focus on Real-Time Compliance

### 2024-08-29 Masterarbeit_Robotik_Pamuk.tex
organize

% Die nachfolgenden Pakete stellen sonst nicht benötigte Features zur Verfügung
\usepackage{blindtext}
%
% Einträge für Deckblatt, Kurzfassung, etc.
%
\title{Virtualisierung eines Echtzeit-Betriebssystems zur Steuerung eines Roboters mit Schwerpunkt auf die
Einhaltung der Echtzeit}
\author{Halil Pamuk, BSc}
\studentnumber{51842568}
%\author{Titel Vorname Name, Titel\and{}Titel Vorname Name, Titel}
%\studentnumber{XXXXXXXXXXXXXXX\and{}XXXXXXXXXXXXXXX}
\supervisor{Sebastian Rauh, MSc. BEng}
%\supervisor[Begutachter]{Titel Vorname Name, Titel}
%\supervisor[Begutachterin]{Titel Vorname Name, Titel}
%\secondsupervisor{Titel Vorname Name, Titel}
%\secondsupervisor[Begutachter]{Titel Vorname Name, Titel}
%\secondsupervisor[Begutachterinnen]{Titel Vorname Name, Titel}
\place{Wien}
\kurzfassung{Virtualisierung von Echtzeitbetriebssystemen für die Robotersteuerung bietet viele Vorteile im Vergleich zu herkömmlichen hardwarebasierten Ansätzen. Lösungen, die auf reiner Hardware basieren, sind oft für einen bestimmten Zweck gebaut und daher in ihrer Flexibilität eingeschränkt. Es ist schwierig, die Hardware-Topologie an neue Anforderungen anzupassen, und es kann kostspielig sein, insbesondere bei der Skalierung von Operationen. Der physische Zugang für Updates und Wartung ist herausfordernd, was zu Ausfallzeiten und Produktivitätsverlusten führt. Während die Virtualisierung diese Probleme adressiert, führt sie zu erhöhtem Overhead und Latenz.

@@

\bigskip \noindent This thesis investigates the virtualization of the proprietory Salamander 4 operating system, using QEMU/KVM. Salamander 4 is built with Yocto and employs hard real-time with Xenomai 3. The primary objective is to bridge the latency gap between the virtualized and bare metal versions in order to ensure deterministic and reliable behavior, which is crucial for real-time robotic applications.

@@

\bigskip \noindent Alltogether, this thesis provides a comprehensive blueprint for making a virtualized guest system real-time capable in a host system with deterministic behavior.}
\keywords{Virtualization, Real-Time Systems, Latency Reduction, Robotic Control}

@@

\noindent Figure~\ref{fig:lstopo} is the output of the~\texttt{lstopo} command and visualizes the hardware nodes of the system, including CPU cores, caches, memory, and I/O devices.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.55\columnwidth]{img/lstopo.png}
	\caption[Host Operating System Hardware Topology]{Host Operating System Hardware Topology}
	\label{fig:lstopo}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\columnwidth]{img/Software-Struktur_einer_LASAL_CPU.png}
	\caption[Structure of Salamander 4 CPU]{Structure of Salamander 4 CPU~\cite{LASALOSSIGMATEK}}
	\label{fig:lasal_cpu}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\columnwidth]{img/RAM_Memory_management.png}
	\caption[Memory Management of Salamander 4]{Memory Management of Salamander 4~\cite{LASALOSSIGMATEK}}
	\label{fig:memory_management}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/kernelshark_combo.png}
	\caption[Guest vCPUs plotted on top of Host Threads in KernelShark]{Guest vCPUs plotted on top of Host Threads in KernelShark}
	\label{fig:kernelshark_combo}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\columnwidth]{img/multiple_varans.png}
	\caption[Synchronization between Systems and their Data Exchange]{Synchronization between Systems and their Data Exchange}
	\label{fig:multiple_varans}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/0hardware/gnuplot_max_latency_hardware.png}
	\caption[Latency Distribution of Salamander 4 Bare Metal]{Latency Distribution of Salamander 4 Bare Metal}
	\label{fig:gnuplot_max_latency_hardware}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/1default/gnuplot_max_latency_default.png}
	\caption[Latency Distribution of Salamander 4 Untuned Virtualization]{Latency Distribution of Salamander 4 Untuned Virtualization}
	\label{fig:gnuplot_max_latency_default}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/01combined/gnuplot_combined_max_latency.png}
	\caption[Initial Comparison of Latency Distribution between Hardware and Virtualization]{Initial Comparison of Latency Distribution between Hardware and Virtualization}
	\label{fig:gnuplot_max_latency_combined}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/2taskset/gnuplot_max_latency_taskset.png}
	\caption[Latency Distribution of Salamander 4 after BIOS Configurations]{Latency Distribution of Salamander 4 Virtualization after tuning BIOS Configurations}
	\label{fig:gnuplot_max_latency_taskset}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/4rt_kernelparam/gnuplot_max_latency_rt_kernelparam.png}
	\caption[Latency Distribution of Salamander 4 after Kernel Configurations]{Latency Distribution of Salamander 4 Virtualization after tuning Kernel Configurations}
	\label{fig:gnuplot_max_latency_rt_kernelparam}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/5rt_kernelparam_host/gnuplot_max_latency_rt_kernelparam_host.png}
	\caption[Latency Distribution of Salamander 4 after Host Configurations]{Latency Distribution of Salamander 4 Virtualization after tuning Host Configurations}
	\label{fig:max_latency_rt_kernelparam_host}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/6rt_kernelparam_host_qemu/gnuplot_max_latency_rt_kernelparam_host_qemu.png}
	\caption[Latency Distribution of Salamander 4 after QEMU Configurations]{Latency Distribution of Salamander 4 Virtualization after tuning QEMU Configurations}
	\label{fig:max_latency_rt_kernelparam_host_qemu}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{masterthesis-documentation/docs/resources/scripts/experiment/generated_latencies/rtos_latencies.png}
	\caption[Robotic Application Latency Values of Salamander 4 Hardware]{Robotic Application Latency Values of Salamander 4 Hardware}
	\label{fig:rtos_latencies}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{masterthesis-documentation/docs/resources/scripts/experiment/generated_latencies/untuned_virt_latencies.png}
	\caption[Robotic Application Latency Values of Salamander 4 Untuned Virtualization]{Robotic Application Latency Values of Salamander 4 Untuned Virtualization}
	\label{fig:untuned_virt_latencies}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{masterthesis-documentation/docs/resources/scripts/experiment/generated_latencies/tuned_virt_latencies.png}
	\caption[Robotic Application Latency Values of Salamander 4 Tuned Virtualization]{Robotic Application Latency Values of Salamander 4 Tuned Virtualization}
	\label{fig:tuned_virt_latencies}
\end{figure}
\begin{table}[H]
	\centering
	\caption{Robotic Application Latency Statistics Tuned Virtualization}
	\label{tab:robotic_application_latency_values_virt_tun}
	\setlength{\tabcolsep}{0.5em} % for the horizontal padding
	{\renewcommand{\arraystretch}{1.2}% for the vertical padding
		\begin{tabular}{|l|l|l|l|l|}
			\hline
			\textbf{Samples} & \textbf{Lat Min (µs)} & \textbf{Lat Avg (µs)} & \textbf{Lat Max (µs)} & \textbf{Std Dev (µs)} \\ \hline
			1000             & 1.219                 & 2.62                  & 3.988            & 0.182                   \\ \hline
		\end{tabular}}
\end{table}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.70\columnwidth]{masterthesis-documentation/docs/sigmatek/xenomai/012456combined/gnuplot_combined_max_latency.png}
	\caption[Comparison of Latency Distribution of Salamander 4 Configurations]{Comparison of Latency Distribution of Salamander 4 under different Configurations}
	\label{fig:max_latency_combined_results}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{masterthesis-documentation/docs/resources/scripts/experiment/generated_latencies/combined_latencies.png}
	\caption[Comparison of Robotic Application Latency of Salamander 4 Configurations]{Comparison of Robotic Application Latency of Salamander 4 under different Configurations}
	\label{fig:combined_latencies}
\end{figure}

>>>>

% Die nachfolgenden Pakete stellen sonst nicht benötigte Features zur Verfügung
\usepackage{blindtext}
%
% Einträge für Deckblatt, Kurzfassung, etc.
%
\title{Virtualization of a Real-Time Operating System for Robot Control with a Focus on Real-Time Compliance}
\author{Halil Pamuk, BSc}
\studentnumber{51842568}
%\author{Titel Vorname Name, Titel\and{}Titel Vorname Name, Titel}
%\studentnumber{XXXXXXXXXXXXXXX\and{}XXXXXXXXXXXXXXX}
\supervisor{Sebastian Rauh, MSc. BEng}
%\supervisor[Begutachter]{Titel Vorname Name, Titel}
%\supervisor[Begutachterin]{Titel Vorname Name, Titel}
%\secondsupervisor{Titel Vorname Name, Titel}
%\secondsupervisor[Begutachter]{Titel Vorname Name, Titel}
%\secondsupervisor[Begutachterinnen]{Titel Vorname Name, Titel}
\place{Wien}
\kurzfassung{Virtualisierung von Echtzeitbetriebssystemen für die Robotersteuerung bietet viele Vorteile im Vergleich zu herkömmlichen hardwarebasierten Ansätzen. Lösungen, die auf reiner Hardware basieren, sind oft für einen bestimmten Zweck gebaut und daher in ihrer Flexibilität eingeschränkt. Es ist schwierig, die Hardware-Topologie an neue Anforderungen anzupassen, und es kann kostspielig sein, insbesondere bei der Skalierung von Operationen. Der physische Zugang für Updates und Wartung ist herausfordernd, was zu Ausfallzeiten und Produktivitätsverlusten führt. Während die Virtualisierung diese Probleme adressiert, führt sie zu erhöhtem Overhead und Latenz.

@@

\bigskip \noindent This thesis investigates the virtualization of the proprietary Salamander 4 operating system, using QEMU/KVM. Salamander 4 is built with Yocto and employs hard real-time with Xenomai 3. The primary objective is to bridge the latency gap between the virtualized and bare metal versions in order to ensure deterministic and reliable behavior, which is crucial for real-time robotic applications.

@@

\bigskip \noindent Altogether, this thesis provides a comprehensive blueprint for making a virtualized guest system real-time capable in a host system with deterministic behavior.}
\keywords{Virtualization, Real-Time Systems, Latency Reduction, Robotic Control}

@@

\noindent Figure~\ref{fig:lstopo} is the output of the~\texttt{lstopo} command and visualizes the hardware nodes of the system, including CPU cores, caches, memory, and I/O devices.
\begin{figure}[H]
	\centering
	\includegraphics[width=0.55\columnwidth]{img/methodology/lstopo.png}
	\caption[Host Operating System Hardware Topology]{Host Operating System Hardware Topology}
	\label{fig:lstopo}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.5\columnwidth]{img/methodology/Software-Struktur_einer_LASAL_CPU.png}
	\caption[Structure of Salamander 4 CPU]{Structure of Salamander 4 CPU~\cite{LASALOSSIGMATEK}}
	\label{fig:lasal_cpu}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\columnwidth]{img/methodology/RAM_Memory_management.png}
	\caption[Memory Management of Salamander 4]{Memory Management of Salamander 4~\cite{LASALOSSIGMATEK}}
	\label{fig:memory_management}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/methodology/kernelshark_combo.png}
	\caption[Guest vCPUs plotted on top of Host Threads in KernelShark]{Guest vCPUs plotted on top of Host Threads in KernelShark}
	\label{fig:kernelshark_combo}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.75\columnwidth]{img/methodology/multiple_varans.png}
	\caption[Synchronization between Systems and their Data Exchange]{Synchronization between Systems and their Data Exchange}
	\label{fig:multiple_varans}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{img/implementation/gnuplot_max_latency_hardware.png}
	\caption[Latency Distribution of Salamander 4 Bare Metal]{Latency Distribution of Salamander 4 Bare Metal}
	\label{fig:gnuplot_max_latency_hardware}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{img/implementation/gnuplot_max_latency_default.png}
	\caption[Latency Distribution of Salamander 4 Untuned Virtualization]{Latency Distribution of Salamander 4 Untuned Virtualization}
	\label{fig:gnuplot_max_latency_default}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{img/implementation/gnuplot_combined_max_latency.png}
	\caption[Initial Comparison of Latency Distribution between Hardware and Virtualization]{Initial Comparison of Latency Distribution between Hardware and Virtualization}
	\label{fig:gnuplot_max_latency_combined}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{img/implementation/gnuplot_max_latency_taskset.png}
	\caption[Latency Distribution of Salamander 4 after BIOS Configurations]{Latency Distribution of Salamander 4 Virtualization after tuning BIOS Configurations}
	\label{fig:gnuplot_max_latency_taskset}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{img/implementation/gnuplot_max_latency_rt_kernelparam.png}
	\caption[Latency Distribution of Salamander 4 after Kernel Configurations]{Latency Distribution of Salamander 4 Virtualization after tuning Kernel Configurations}
	\label{fig:gnuplot_max_latency_rt_kernelparam}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{img/implementation/gnuplot_max_latency_rt_kernelparam_host.png}
	\caption[Latency Distribution of Salamander 4 after Host Configurations]{Latency Distribution of Salamander 4 Virtualization after tuning Host Configurations}
	\label{fig:max_latency_rt_kernelparam_host}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\columnwidth]{img/implementation/gnuplot_max_latency_rt_kernelparam_host_qemu.png}
	\caption[Latency Distribution of Salamander 4 after QEMU Configurations]{Latency Distribution of Salamander 4 Virtualization after tuning QEMU Configurations}
	\label{fig:max_latency_rt_kernelparam_host_qemu}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/implementation/rtos_latencies.png}
	\caption[Robotic Application Latency Values of Salamander 4 Hardware]{Robotic Application Latency Values of Salamander 4 Hardware}
	\label{fig:rtos_latencies}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/implementation/untuned_virt_latencies.png}
	\caption[Robotic Application Latency Values of Salamander 4 Untuned Virtualization]{Robotic Application Latency Values of Salamander 4 Untuned Virtualization}
	\label{fig:untuned_virt_latencies}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/implementation/tuned_virt_latencies.png}
	\caption[Robotic Application Latency Values of Salamander 4 Tuned Virtualization]{Robotic Application Latency Values of Salamander 4 Tuned Virtualization}
	\label{fig:tuned_virt_latencies}
\end{figure}
\begin{table}[H]
	\centering
	\caption{Robotic Application Latency Statistics Tuned Virtualization}
	\label{tab:robotic_application_latency_values_virt_tun}
	\setlength{\tabcolsep}{0.5em} % for the horizontal padding
	{\renewcommand{\arraystretch}{1.2}% for the vertical padding
		\begin{tabular}{|l|l|l|l|l|}
			\hline
			\textbf{Samples} & \textbf{Lat Min (µs)} & \textbf{Lat Avg (µs)} & \textbf{Lat Max (µs)} & \textbf{Std Dev (µs)} \\ \hline
			1000             & 1.219                 & 2.62                  & 3.988            & 0.182                   \\ \hline
		\end{tabular}}
\end{table}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.70\columnwidth]{img/results/gnuplot_combined_max_latency_all.png}
	\caption[Comparison of Latency Distribution of Salamander 4 Configurations]{Comparison of Latency Distribution of Salamander 4 under different Configurations}
	\label{fig:max_latency_combined_results}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=1.0\columnwidth]{img/results/combined_latencies.png}
	\caption[Comparison of Robotic Application Latency of Salamander 4 Configurations]{Comparison of Robotic Application Latency of Salamander 4 under different Configurations}
	\label{fig:combined_latencies}
\end{figure}
